---
title: "Quality control and normalisation"
author: "ejjohnson93"
date: "2022-09-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Load libraries

```{r, message=FALSE, warning=FALSE}

library(tidyverse)
library(NormalyzerDE) # compare normalisation strategies 
library(pheatmap) # visualise whether samples cluster by missingness 
library(cowplot) # pretty theme 
library(ggpubr) # utilities to arrange/annotate ggplot2 plots
library(proDA) # median normalisation function
library(HybridMTest) # per-row Pearson corrrelation 
library(SummarizedExperiment)
library(DEP)
source("./code/functions.R") # load helper functions 

```


# Proteomics 

## Load data

Read in data and accompanying phenotype tables: 

```{r}

# Data
plasma_data <- read.csv("./data/plasma_proteins.csv", row.names = 1) 
colnames(plasma_data) <- gsub(pattern = "X", replacement = "Sample_", x = colnames(plasma_data))
  
sf_data <- read.csv("./data/sf_proteins.csv", row.names = 1)
colnames(sf_data) <- gsub(pattern = "X", replacement = "Sample_", x = colnames(sf_data))

# Pheno
plasma_pheno <- read.csv("./data/plasma_pheno.csv")
sf_pheno <- read.csv("./data/sf_pheno.csv")

# Annotation
plasma_annotation <- read.csv("./data/plasma_protein_annotation.csv", row.names=1)
sf_annotation <- read.csv("./data/sf_protein_annotation.csv", row.names=1)

```


Process phenotype table and make sure it is in the same order as the raw data: 

```{r}

plasma_pheno$Sample.Number <- paste0("Sample_", plasma_pheno$Sample.Number)
plasma_pheno$Horse <- as.factor(plasma_pheno$Horse)
plasma_pheno <- plasma_pheno[match(colnames(plasma_data),plasma_pheno$Sample.Number),]


sf_pheno$Sample.Number <- paste0("Sample_", sf_pheno$Sample.Number)
sf_pheno$Horse <- as.factor(sf_pheno$Horse)
sf_pheno <- sf_pheno[match(colnames(sf_data),sf_pheno$Sample.Number),]

```


## Normalisation

To compare normalisation strategies the NormalyzerDE package was used. This requires a design matrix and a summarised experimental object to be constructed. The design matrix needs to contain a column with a 'group' header. The summarised experimental object won't accept values <1 as the data is log-transformed during processing this will lead to negative values which in turn will crash processing. For the purpose of comparing normalisation techniques a small constant can be added to each value (in this case 1) before normalisation (not including NA values). 

```{r, eval=FALSE}

# Create directory for outputs
if (!dir.exists("./normalisation")) {
    dir.create("./normalisation")
}


# Create design matrix
design <- plasma_pheno[,c(5,2,1)]
colnames(design) <- c("sample", "group", "batch")


# Create a summarised experiment object
sumExpObj <- SummarizedExperiment::SummarizedExperiment(
as.matrix(plasma_data +1), # Add a constant to each value
colData=design,
rowData=rownames(data)
)

# Run analysis 
normalyzer(jobName="proteomics_norm", experimentObj = sumExpObj, outputDir=dir)

```


Box plots of some of the normalisation methods: 

```{r}

# Read in sample normalised data 
vsn_normalised <- read.delim("./normalisation/proteomics_norm/VSN-normalized.txt", row.names = 1)
log2_normalised <- read.delim("./normalisation/proteomics_norm/log2-normalized.txt", row.names = 1)
median_normalised <- read.delim("./normalisation/proteomics_norm/median-normalized.txt", row.names = 1)

```

```{r}

# Boxplots of VSN, median and log2 normalised data
boxplot(vsn_normalised)
boxplot(log2_normalised)
boxplot(median_normalised)

```


Log2 transform the data:

```{r}

sf_data_norm <- log2(sf_data) 
plasma_data_norm <- log2(plasma_data) 

# Median normalisation would be:
#sf_data_norm <- log2(sf_data) %>% as.matrix(.) %>% median_normalization(.)
#plasma_data_norm <- log2(plasma_data) %>% as.matrix(.) %>% median_normalization(.)

```

Note: median normalisation was also tested for EDA and DE analysis.


## Quality check 

Create box plots of the log transformed data:

```{r}

sf_bp <- t(sf_data_norm) %>%
  as.data.frame() %>%
  mutate(Day = as.factor(sf_pheno$Day), Horse = sf_pheno$Horse, Group = sf_pheno$Group) %>%
  rownames_to_column('Sample_ID') %>%
  pivot_longer(cols = A0A061DBP6:Q9XSZ5) %>%
  ggplot(aes(x=Sample_ID, y=value, fill=Day)) + geom_boxplot() + theme_cowplot() +
    theme(axis.text.x = element_text(angle = 90, size = 8)) + 
    xlab("Sample ID") + 
    ylab("Log2 Intensity") + 
    ggtitle("Synovial fluid EVs")


```

```{r}

plasma_bp <- t(plasma_data_norm) %>%
  as.data.frame() %>%
  mutate(Day = as.factor(plasma_pheno$Day), Horse = plasma_pheno$Horse, Group = plasma_pheno$Group) %>%
  rownames_to_column('Sample_ID') %>%
  pivot_longer(cols = A0A061DBP6:Q9GKX7) %>%
  ggplot(aes(x=Sample_ID, y=value, fill=Day)) + geom_boxplot() + theme_cowplot() +
    theme(axis.text.x = element_text(angle = 90, size = 8)) + 
    xlab("Sample ID") + 
    ylab("Log2 Intensity") + 
    ggtitle("Plasma EVs")

```


```{r}

sf_missing_values <- data.frame(missing_count = colSums(is.na(sf_data_norm)),
                    Day = as.factor(sf_pheno$Day),
                    Horse = sf_pheno$Horse) %>%
                    rownames_to_column('Sample_ID') 


sf_missing_values_barplot <- ggbarplot(sf_missing_values, x = "Sample_ID", y = "missing_count",  fill = "Day", scales = "free") + 
    #scale_fill_brewer(palette="Set1") + 
    theme_pubclean() +
    theme(legend.position="right") + theme_cowplot() +
    theme(axis.text.x = element_text(angle = 90, size = 8)) + 
    xlab("Sample ID") + 
    ylab("# Missing values") 


```

```{r}

plasma_missing_values <- data.frame(missing_count = colSums(is.na(plasma_data_norm)),
                    Day = as.factor(plasma_pheno$Day),
                    Horse = plasma_pheno$Horse) %>%
                    rownames_to_column('Sample_ID') 


plasma_missing_values_barplot <- ggbarplot(plasma_missing_values, x = "Sample_ID", y = "missing_count",  fill = "Day", scales = "free") + 
    #scale_fill_brewer(palette="Set1") + 
    theme_pubclean() +
    theme(legend.position="right") + theme_cowplot() +
    theme(axis.text.x = element_text(angle = 90, size = 8)) + 
    xlab("Sample ID") + 
    ylab("# Missing values") 


```


```{r, echo=FALSE, fig.height = 16, fig.width = 7, fig.align="center"}

ggarrange(sf_bp, sf_missing_values_barplot, plasma_bp, plasma_missing_values_barplot,                        
          nrow = 4, 
          ncol = 1,
          labels = c("A", "B", "C", "D")                                       
          ) 

```


For the synovial fluid samples, samples 6, 13, 17, 43 and 44 have a much lower number of missing values, this could correspond to a batch effect. 

Its also notable that the intensity distribution is generally shifted upwards for samples that have more missing values. This agrees with the idea that small values are more likely to be missing. 



## Missing values - heat maps and correlation

### Synovial fluid samples

An additional way to analyse the missing values is to replace them all with 0's and then carry out clustering analysis. The rationale behind this is that if the missing values correspond to a biological effect (e.g. proteins that were abundant at time 0 that are absent at later time points or vice versa) they will cluster according to that biological effect. 

Firstly, all the NA's in the matrix need to be replaced with zeros: 

```{r}

# Create a copy of the data
sf_heatmap_data <- sf_data_norm

# Replace all the NA values in the copy with 0's
sf_heatmap_data[is.na(sf_heatmap_data)] <- 0

```


Next we need to create a dataframe that contains the annotation for all the features of interest in the phenotype table. For the synovial fluid samples this is day, group and horse. 

```{r}

my_sample_col <- data.frame(Horse = sf_pheno$Horse, Day = sf_pheno$Day, Group = sf_pheno$Group)
rownames(my_sample_col) <- sf_pheno$Sample.Number

```


Next create a heatmap using the pheatmap package taking the modified data and annotation as input. Multiple clustering methods were used, but they all resulted in a similar pattern.

```{r}

sf_heatmap <- pheatmap(sf_heatmap_data, 
                       annotation_col = my_sample_col, 
                       show_rownames=F,
                       cluster_cols = TRUE,
                       clustering_method = "ward.D2"
                       )

# Use the helper function 'save_pheatmap' to write the pheatmap output ot a figure
save_pheatmap_png(sf_heatmap, "./output/sf_missing_values_heatmap.png")

```


For the synovial fluid samples the missing values don't appear to be entirely at random. Horse 4 seems to display some clustering, the samples that have fewer missing values cluster together (6, 13, 17, 43 and 44). This may correspond to a batch effect in the processing. 

We can also create a clustering dendogram to show just the clustering:

```{r}

my_hclust_sample <- hclust(dist(t(sf_heatmap_data)), method = "complete")

as.dendrogram(my_hclust_sample) %>%
  plot()

```

It is better to avoid imputing missing values in proteomics data, especially when there are a low number of replicates per group as you might end up imputing based off one value and this will result in unreliable data and potential false positives. Ideally only the proteins without missing values should be taken forward for differential expression analysis. However, the synovial fluid data is very sparse: 

```{r}

dim(sf_data_norm[which(rowMeans(is.na(sf_data_norm)) < 0.1), ])
dim(na.omit(sf_data_norm))

```

There are only 130 proteins remaining if we take rows with <10% missing values and 27 proteins remaining if we take rows that have no missing values. 


### Plasma samples

Now repeat the same process for the plasma samples: 

```{r}

plasma_heatmap_data <- plasma_data_norm
plasma_heatmap_data[is.na(plasma_heatmap_data)] <- 0

```

```{r}

my_sample_col <- data.frame(Horse = plasma_pheno$Horse, Day = plasma_pheno$Day)
rownames(my_sample_col) <- plasma_pheno$Sample.Number

plasma_heatmap <- pheatmap(plasma_heatmap_data, annotation_col = my_sample_col, show_rownames=F)

save_pheatmap_png(plasma_heatmap, "./output/plasma_missing_values_heatmap.png")

```

There appears to be some clustering dependent on which horse the plasma was extracted from. Additionally, some mild clustering based off the time point. 

```{r}

my_hclust_sample <- hclust(dist(t(plasma_heatmap_data)), method = "complete")

as.dendrogram(my_hclust_sample) %>%
  plot()

```



## Missing values - additional analysis and quality control

### Synovial fluid samples

The synovial fluid data was subset to only include the OA time course. This is so the DEP package filters the samples in a treatment dependent manner. 

```{r}

sf_pheno_OA <- sf_pheno[!sf_pheno$Group == "Control",]
sf_data_norm_OA <- sf_data_norm[,sf_pheno_OA$Sample.Number]

```

#### DEP package tools for analysing missing values

```{r}

sf_design <- sf_pheno_OA[,c(6,6,2,1)]
colnames(sf_design) <- c("label", "ID", "condition", "replicate")
sf_design$replicate <- as.numeric(sf_design$replicate)
sf_row_dat <- cbind(data.frame(ID = c(1:1201)), sf_data_norm_OA) %>% rownames_to_column(., "name")


sf_se <- SummarizedExperiment(assays=sf_data_norm_OA,
                                  rowData=sf_row_dat, colData=sf_design)


```

```{r}

plot_frequency(sf_se)

```

```{r}

sf_se_filter <- filter_proteins(sf_se, "condition", thr = 0)

```

```{r}

plot_missval(sf_se)
plot_missval(sf_se_filter)

```

Samples 4, 15, 48, 39 all appear to be very poor quality. Might be best to remove them from the analysis altogether. However, removing 4, 15 and 39 would mean only two replicates for the 49 day and 63 day time points.


```{r}

plot_detect(sf_se)
plot_detect(sf_se_filter)


```


Subset normalised data by filtered missing values: 

```{r}

sf_data_norm <- sf_data_norm[sf_se_filter@NAMES,]

```


### Plasma samples

#### Correlation between missing values and time - Pearson's correlation

```{r}

plasma_data_norm_missing <- plasma_data_norm %>%
  mutate_if(is.numeric, ~1 * (. > 0)) %>% # Replace numeric values with 1's
  replace(is.na(.), 0) # Replace NA's with 0's



plasma_data_norm_missing <- plasma_data_norm_missing %>% 
  mutate("Day_0" = rowSums(plasma_data_norm_missing[,plasma_pheno$Day == "0"]),
         "Day_10" = rowSums(plasma_data_norm_missing[,plasma_pheno$Day == "10"]),
         "Day_35" = rowSums(plasma_data_norm_missing[,plasma_pheno$Day == "35"]),
         "Day_42" = rowSums(plasma_data_norm_missing[,plasma_pheno$Day == "42"]),
         "Day_49" = rowSums(plasma_data_norm_missing[,plasma_pheno$Day == "49"]),
         "Day_56" = rowSums(plasma_data_norm_missing[,plasma_pheno$Day == "56"]),
         "Day_63" = rowSums(plasma_data_norm_missing[,plasma_pheno$Day == "63"])) 

```

```{r}

plasma_data_norm_missing <- as.matrix(plasma_data_norm_missing[,27:33])

```

```{r}

time_points <- c(0, 10, 35, 42, 49, 56, 63)
pearson <- row.pearson(plasma_data_norm_missing, time_points)

```


#### DEP package tools for analysing missing values

```{r}

plasma_design <- plasma_pheno[,c(5,5,2,1)]
colnames(plasma_design) <- c("label", "ID", "condition", "replicate")
plasma_design$replicate <- as.numeric(plasma_design$replicate)
plasma_row_dat <- cbind(data.frame(ID = c(1:259)), plasma_data_norm) %>% rownames_to_column(., "name")


plasma_se <- SummarizedExperiment(assays=plasma_data_norm,
                     rowData=plasma_row_dat, colData=plasma_design)


```

```{r}

plot_frequency(plasma_se)

```

```{r}

plasma_se_filter <- filter_proteins(plasma_se, "condition", thr = 0)

```

```{r}

plot_missval(plasma_se)
plot_missval(plasma_se_filter)

```

```{r}

plot_detect(plasma_se)
plot_detect(plasma_se_filter)


```


Subset normalised data by filtered missing values: 

```{r}

plasma_data_norm <- plasma_data_norm[plasma_se_filter@NAMES,]

```

## Write data

Save the R data objects for use in exploratory analysis and differential expression analysis. 

```{r}

# Plasma 
saveRDS(plasma_data_norm, "./data/plasma_data_norm.rds")
saveRDS(plasma_pheno, "./data/plasma_pheno.rds")
saveRDS(plasma_annotation, "./data/plasma_proteomics_annotation.rds")

# SF 
saveRDS(sf_data_norm, "./data/sf_data_norm.rds")
saveRDS(sf_pheno, "./data/sf_pheno.rds")
saveRDS(sf_annotation, "./data/sf_proteomics_annotation.rds")

```



# Sequencing data 

```{r}



```




