---
title: "Differential expression analysis"
author: "ejjohnson93"
date: "2022-09-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Load libraries

```{r, message=FALSE, warning=FALSE}

library(tidyverse)
library(proDA)
library(DT)
library(openxlsx)
library(limma)

```

# With missing values

## Load data


```{r}

# Data
sf_data_norm <- readRDS("./data/sf_data_norm.rds")
plasma_data_norm <- readRDS("./data/plasma_data_norm.rds")


# Pheno
sf_pheno <- readRDS("./data/sf_pheno.rds")
plasma_pheno <- readRDS("./data/plasma_pheno.rds")

```


## Plasma DE analysis 

### Version 1, time as a factor 

```{r}

plasma_design <- plasma_pheno[,c(5,2,1)]
colnames(plasma_design) <- c("name", "day", "horse")
plasma_design$day <- as.factor(plasma_design$day)

```



```{r}

fit <- proDA(as.matrix(plasma_data_norm), design = ~ day + horse, 
             col_data = plasma_design, reference_level = "0") # Manually specified reference level but confirmed its 0 by default

```


`result_names(fit)` can be used to view the names of the possible contrasts. 

```{r}

plasma_day63 <- test_diff(fit, day63) %>% arrange(-desc(adj_pval))
plasma_day56 <- test_diff(fit, day56) %>% arrange(-desc(adj_pval))
plasma_day49 <- test_diff(fit, day49) %>% arrange(-desc(adj_pval))
plasma_day42 <- test_diff(fit, day42) %>% arrange(-desc(adj_pval))
plasma_day35 <- test_diff(fit, day35) %>% arrange(-desc(adj_pval))
plasma_day10 <- test_diff(fit, day10) %>% arrange(-desc(adj_pval))

```

Day 35 and 42 have one sig protein after adjustment each: A0A3Q2HID2 & 	A0A3Q2H659.  

These results were combined into a list and then saved as an Excel spreadsheet: 

```{r}

plasma_DE_res <- list("day63" = plasma_day63, "day56" = plasma_day56, "day49" = plasma_day49, "day42" = plasma_day42, "day35" = plasma_day35, "day10" = plasma_day10)
openxlsx::write.xlsx(plasma_DE_res, file = "./output/plasma_proDA_DE_results.xlsx")

```


Sample results showing the day 42 vs day 0 contrast are below. The rows are ordered by descending p-value. 

```{r, echo=FALSE}

DT::datatable(plasma_day42[,c(1:4,6,10)], 
              class = "cell-border stripe", 
              filter= "top", 
              extensions = "Buttons",
              options = list(pageLength = 10, pageWidth = 5, dom = 'Bfrtip',
              buttons = c('copy', 'print'))
              )

```



### Version 2, time as a continuous variable 

```{r}

plasma_design <- plasma_pheno[,c(5,2,1)]
colnames(plasma_design) <- c("name", "day", "horse")

```



```{r}

fit <- proDA(as.matrix(plasma_data_norm), design = ~ day + horse, 
             col_data = plasma_design) 

```



```{r}

plasma_time <- test_diff(fit, day) %>% arrange(-desc(adj_pval))

```



## Synovial fluid differential expression analysis 

### Version 1, time and condition as factors 

```{r}

sf_design <- sf_pheno[,c(6,2,1,5)]
colnames(sf_design) <- c("name", "day", "horse", "group")
sf_design$group_day <- as.factor(paste0(sf_design$group, "_", sf_design$day))

```



```{r}

fit <- proDA(as.matrix(sf_data_norm), design = ~ group_day + horse, 
             col_data = sf_design) # Manually specified reference level but confirmed its 0 by default

```


`result_names(fit)` can be used to view the names of the possible contrasts. Create contrasts between all the OA samples vs time 0 and sort them by descending p-value. 

```{r}

sf_day63 <- test_diff(fit, group_dayOA_63 - group_dayOA_0) %>% arrange(-desc(pval))
sf_day56 <- test_diff(fit, group_dayOA_56 - group_dayOA_0) %>% arrange(-desc(pval))
sf_day49 <- test_diff(fit, group_dayOA_49 - group_dayOA_0) %>% arrange(-desc(pval))
sf_day42 <- test_diff(fit, group_dayOA_42 - group_dayOA_0) %>% arrange(-desc(pval))
sf_day35 <- test_diff(fit, group_dayOA_35 - group_dayOA_0) %>% arrange(-desc(pval))
sf_day10 <- test_diff(fit, group_dayOA_10 - group_dayOA_0) %>% arrange(-desc(pval))

```


These results can be combined into a list and then saved as an Excel spreadsheet: 

```{r}

sf_DE_res <- list("day63" = sf_day63, "day56" = sf_day56, "day49" = sf_day49, "day42" = sf_day42, "day35" = sf_day35, "day10" = sf_day10)
openxlsx::write.xlsx(sf_DE_res, file = "./output/sf_proDA_DE_results.xlsx")

```


Sample results of the differential expression analysis with the time points as factors are shown below. The results shown are day 63 versus day 0. The rows are ordered by descending p-value. 

```{r, echo=FALSE}

DT::datatable(sf_day63[,c(1:4,6,10)], 
              class = "cell-border stripe", 
              filter= "top", 
              extensions = "Buttons",
              options = list(pageLength = 10, pageWidth = 5, dom = 'Bfrtip',
              buttons = c('copy', 'print'))
              )

```


Note that when median normalisation is used instead there are two differentially expressed proteins at day 10 versus day 0: F6PH38 & A0A5F5PPB8. However, when this noramalisation technique is used the plasma proteins no longer pass FDR.


### Version 2, time as a continuous variable 

```{r}

sf_pheno_OA <- sf_pheno[!sf_pheno$Group == "Control",]
sf_data_norm_just_OA <- sf_data_norm[,sf_pheno_OA$Sample.Number]

```

```{r}

sf_design <- sf_pheno_OA[,c(6,2,1)]
colnames(sf_design) <- c("name", "day", "horse")

```


```{r}

fit <- proDA(as.matrix(sf_data_norm_just_OA), design = ~ day + horse, 
             col_data = sf_design) 

```


```{r}

sf_time <- test_diff(fit, day) %>% arrange(-desc(pval))

```


Results of the differential expression analysis with time as a continuous variable are shown below. The rows are ordered by descending p-value. 

```{r, echo=FALSE}

DT::datatable(sf_time[,c(1:4,6,10)], 
              class = "cell-border stripe", 
              filter= "top", 
              extensions = "Buttons",
              options = list(pageLength = 10, pageWidth = 5, dom = 'Bfrtip',
              buttons = c('copy', 'print'))
              )

```



# Without missing values

## Load and format data

```{r}

# Data
sf_data_norm <- readRDS("./data/sf_data_norm.rds")
plasma_data_norm <- readRDS("./data/plasma_data_norm.rds")


# Pheno
sf_pheno <- readRDS("./data/sf_pheno.rds")
plasma_pheno <- readRDS("./data/plasma_pheno.rds")

```


Remove all the rows with NA values: 

```{r}

sf_data_norm <- na.omit(sf_data_norm)
plasma_data_norm <- na.omit(plasma_data_norm)

#plasma_data_norm <- plasma_data_norm[which(rowMeans(is.na(plasma_data_norm)) < 0.1), ]

```


Remove the control samples for the synovial fluid. For both plasma and synovial fluid DE relative to time zero will be investigated. 

```{r}

sf_pheno <- sf_pheno[!sf_pheno$Group == "Control",]
sf_data_norm <- sf_data_norm[,sf_pheno$Sample.Number]

```


## Plasma DE analysis 

### Version 1, time as a factor 


```{r}

# Create the model matrix using this factor
design <- model.matrix(~0 + as.factor(plasma_pheno$Day) + as.factor(plasma_pheno$Horse))

# Rename the columns of the design matrix using the levels of the factor 
colnames(design) <- c("Day_0", "Day_10", "Day_35", "Day_42", "Day_49", "Day_56", "Day_63", "Horse_4", "Horse_6", "Horse_8")


```

```{r}

# Create contrast matrix 
contrast <- makeContrasts(day_63vs0 = Day_63 - Day_0,
                          day_56vs0 = Day_56 - Day_0,
                          day_49vs0 = Day_49 - Day_0,
                          day_42vs0 = Day_42 - Day_0,
                          day_35vs0 = Day_35 - Day_0,
                          day_10vs0 = Day_10 - Day_0,
                          levels = design)

# Fit model
fit <- lmFit(plasma_data_norm, design)
fit2 <- contrasts.fit(fit, contrast)
fit2 <- eBayes(fit2)

# Number of differentially expressed proteins at each time point 
summary(decideTests(fit2))

```

```{r}

plasma_tt_day56 <- topTable(fit2, coef=2, n=Inf)
plasma_tt_day35 <- topTable(fit2, coef=5, n=Inf)

```

[A0A3Q2HID2](https://www.uniprot.org/uniprotkb/A0A3Q2HID2/entry) (Coagulation factor XIII A chain) is the one significant protein at both time points.


```{r, echo=FALSE}

DT::datatable(plasma_tt_day56, 
              class = "cell-border stripe", 
              filter= "top", 
              extensions = "Buttons",
              options = list(pageLength = 10, pageWidth = 5, dom = 'Bfrtip',
              buttons = c('copy', 'print'))
              )

```


### Version 2, time as continuous variable

```{r}

# Create the model matrix using this factor
design <- model.matrix(~ plasma_pheno$Day + as.factor(plasma_pheno$Horse))

# Rename the columns of the design matrix using the levels of the factor 
colnames(design) <- c("(Intercept", "Day", "Horse_4", "Horse_6", "Horse_8")


```

```{r}

# Fit model
fit <- lmFit(plasma_data_norm, design)
fit2 <- eBayes(fit)

# Number of differentially expressed genes
summary(decideTests(fit2))

```

```{r}

plasma_tt_continuous_time <- topTable(fit2, coef=2, n=Inf)

```

## Synovial fluid DE analysis 

### Version 1, time as a factor 

```{r}

# Create the model matrix using this factor
design <- model.matrix(~0 + as.factor(sf_pheno$Day) + as.factor(sf_pheno$Horse))

# Rename the columns of the design matrix using the levels of the factor 
colnames(design) <- c("Day_0", "Day_10", "Day_35", "Day_42", "Day_49", "Day_56", "Day_63", "Horse_4", "Horse_6", "Horse_8")


```

```{r}

# Create contrast matrix 
contrast <- makeContrasts(day_63vs0 = Day_63 - Day_0,
                          day_56vs0 = Day_56 - Day_0,
                          day_49vs0 = Day_49 - Day_0,
                          day_42vs0 = Day_42 - Day_0,
                          day_35vs0 = Day_35 - Day_0,
                          day_10vs0 = Day_10 - Day_0,
                          levels = design)

# Fit model
fit <- lmFit(sf_data_norm, design)
fit2 <- contrasts.fit(fit, contrast)
fit2 <- eBayes(fit2)

# Number of differentially expressed proteins at each time point 
summary(decideTests(fit2))

```

```{r}

sf_tt_day10 <- topTable(fit2, coef=6, n=Inf)

```

```{r, echo=FALSE}

DT::datatable(sf_tt_day10, 
              class = "cell-border stripe", 
              filter= "top", 
              extensions = "Buttons",
              options = list(pageLength = 10, pageWidth = 5, dom = 'Bfrtip',
              buttons = c('copy', 'print'))
              )

```

### Version 2, time as continuous variable

```{r}

# Create the model matrix using this factor
design <- model.matrix(~ sf_pheno$Day + as.factor(sf_pheno$Horse))

# Rename the columns of the design matrix using the levels of the factor 
colnames(design) <- c("(Intercept", "Day", "Horse_4", "Horse_6", "Horse_8")


```

```{r}

# Fit model
fit <- lmFit(sf_data_norm, design)
fit2 <- eBayes(fit)

# Number of differentially expressed genes
summary(decideTests(fit2))

```

Time as a continuous variable results in no significant proteins (suggests the response must be non-linear).


