---
title: "QC & normalisation - sequencing data"
author: "ejjohnson93"
date: "2022-09-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
bibliography: references.bib
---

# Load libraries

```{r, message=FALSE, warning=FALSE}

source("./code/utility-functions.R") # load helper functions 
library(tidyverse)
library(edgeR)
library(cowplot) # ggplot2 theme
library(viridis) # colour schemes
library(ggpubr) # utilities to arrange/annotate ggplot2 plots
library(R.utils) # function to count zero values
library(PCAtools) # pca & correlation plot

```

# Phenotype

First load the phenotype and sequencing info data:

```{r}

# Pheno
plasma_pheno <- read.csv("./data/plasma_pheno.csv")
sf_pheno <- read.csv("./data/sf_pheno.csv")


# Sequencing info
plasma_sequencing_info <- read.csv("./data/plasma_sequencing_info.csv")
sf_sequencing_info <- read.csv("./data/sf_sequencing_info.csv")


```

Process the phenotype and sequencing info tables and combine them:

```{r}

# Match pheno data to sequencing info
plasma_pheno <- plasma_pheno[match(plasma_sequencing_info$customerID, plasma_pheno$Sample.Number),]
sf_pheno <- sf_pheno[match(sf_sequencing_info$customerID, sf_pheno$Sample.Number),]


# Convert sequencing info to numeric
#plasma_sequencing_info[, 3:17] <- sapply(plasma_sequencing_info[, 3:17], as.numeric)
#sf_sequencing_info[, 3:17] <- sapply(sf_sequencing_info[, 3:17], as.numeric)


# Join sequencing and pheno information
plasma_pheno <- cbind(plasma_pheno, plasma_sequencing_info)
sf_pheno <- cbind(sf_pheno, sf_sequencing_info)


# Process phenotype data to include a 'sample_' prefix in front of the sample IDs
# Convert horse ID to a factor
plasma_pheno$Sample.Number <- paste0("Sample_", plasma_pheno$Sample.Number)
plasma_pheno$Horse <- as.factor(plasma_pheno$Horse)

sf_pheno$Sample.Number <- paste0("Sample_", sf_pheno$Sample.Number)
sf_pheno$Horse <- as.factor(sf_pheno$Horse)
sf_pheno <- sf_pheno[,-8]

```

Next, load and process each type of sequencing data.

# miRNA

The data was provided in CPM format. The reports TAmiRNA provided contain the raw counts for the miRNA. To assess the normalisation the raw counts and the CPM normalised counts were both read in.

## Plasma

### Read data in

```{r}

# Raw miRNA counts
plasma_miRNA_data_raw <- read.csv("./data/raw_miRNA_plasma.csv", row.names = 1) 
colnames(plasma_miRNA_data_raw) <- gsub(pattern = "X", replacement = "Sample_", x = colnames(plasma_miRNA_data_raw))


# CPM normalised plasma miRNA data
plasma_miRNA_data <- read.csv("./data/miRNA_plasma.csv", row.names = 1)
colnames(plasma_miRNA_data) <- gsub(pattern = "X", replacement = "Sample_", x = colnames(plasma_miRNA_data))

```

Process raw data and make sure its in the same order as the phenotype table:

```{r}

plasma_miRNA_data <- plasma_miRNA_data[,match(plasma_pheno$Sample.Number, colnames(plasma_miRNA_data))]

```

### Normalisation

Manually applied TMM normalisation then converted the counts into CPM.

```{r}

plasma_miRNA_data_DGE <- DGEList(counts = plasma_miRNA_data_raw, lib.size = colSums(plasma_miRNA_data), samples = plasma_pheno)
plasma_miRNA_data_DGE_TMM <- calcNormFactors(plasma_miRNA_data_DGE, method = "TMM")

raw_counts_CPM <- cpm(plasma_miRNA_data_DGE, log=FALSE)
raw_counts_CPM_TMM <- cpm(plasma_miRNA_data_DGE_TMM, log=FALSE)

```

### Plot data

Compared the distribution of the data provided to the TMM normalised CPM values.

```{r, message=FALSE}

bp_plasma_miRNA_data <- gg_boxplot(plasma_miRNA_data,
                                   log.transform=T,
                                   title="Plasma miRNA - original CPM data provided",
                                   xlab="Sample ID", ylab="Log2 CPM")



bp_raw_counts_CPM_TMM <- gg_boxplot(raw_counts_CPM_TMM,
                                   log.transform=T,
                                   title="Plasma miRNA - raw counts, TMM and CPM normalised",
                                   xlab="Sample ID", ylab="Log2 CPM")



bp_plasma_miRNA_data_raw <- gg_boxplot(plasma_miRNA_data_raw,
                                   log.transform=T,
                                   title="Plasma miRNA - Log2 raw counts",
                                   xlab="Sample ID", ylab="Log2 CPM")

  

bp_raw_counts_CPM <- gg_boxplot(raw_counts_CPM,
                                   log.transform=T,
                                   title="Plasma miRNA - raw counts, just CPM normalised",
                                   xlab="Sample ID", ylab="Log2 CPM")

  

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

bp_plasma_miRNA_data
bp_raw_counts_CPM_TMM
bp_plasma_miRNA_data_raw
bp_raw_counts_CPM

```

## Synovial fluid

### Read data in

```{r}

# CPM normalised SF miRNA data
sf_miRNA_data <- read.csv("./data/miRNA_sf.csv", row.names = 1)
colnames(sf_miRNA_data) <- gsub(pattern = "X", replacement = "Sample_", x = colnames(sf_miRNA_data))

```

Process raw data and make sure its in the same order as the phenotype table:

```{r}

sf_miRNA_data <- sf_miRNA_data[,match(sf_pheno$Sample.Number, colnames(sf_miRNA_data))]

```

### Plot data

Look at the distribution of the data provided...

```{r, message=FALSE}

bp_sf_miRNA_data <- gg_boxplot(sf_miRNA_data,
                                   log.transform=T,
                                   title="Synovial fluid miRNA - original CPM data provided",
                                   xlab="Sample ID", ylab="Log2 CPM")

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

bp_sf_miRNA_data

```

# lncRNA

The rest of the sequencing data didn't have raw counts.The CPM data provided will be read in for the lncRNA, snoRNA, snRNA, tRNA.

## Plasma:

### Read data in

```{r}

# CPM normalised plasma lncRNA data
plasma_lncRNA_data <- read.csv("./data/lncRNA_plasma.csv", row.names = 1)
colnames(plasma_lncRNA_data) <- gsub(pattern = "X", replacement = "Sample_", x = colnames(plasma_lncRNA_data))

```

Process raw data and make sure its in the same order as the phenotype table:

```{r}

plasma_lncRNA_data <- plasma_lncRNA_data[,match(plasma_pheno$Sample.Number, colnames(plasma_lncRNA_data))]

```

Look at the distribution of the data provided...

```{r, message=FALSE}

bp_plasma_lncRNA_data <- gg_boxplot(plasma_lncRNA_data,
                                log.transform =T,
                                title="Plasma lncRNA",
                                xlab="Sample ID", ylab="Log2 CPM")

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

bp_plasma_lncRNA_data

```

## Synovial fluid:

### Read data in

```{r}

# CPM normalised SF lncRNA data
sf_lncRNA_data <- read.csv("./data/lncRNA_sf.csv", row.names = 1)
colnames(sf_lncRNA_data) <- gsub(pattern = "X", replacement = "Sample_", x = colnames(sf_lncRNA_data))

```

Process raw data and make sure its in the same order as the phenotype table:

```{r}

sf_lncRNA_data <- sf_lncRNA_data[,match(sf_pheno$Sample.Number, colnames(sf_lncRNA_data))]

```

Look at the distribution of the data provided...

```{r, message=FALSE}

bp_sf_lncRNA_data <- gg_boxplot(sf_lncRNA_data,
                               log.transform=T,
                               title="Synovial fluid lncRNA",
                               xlab="Sample ID", ylab="Log2 CPM")

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

bp_sf_lncRNA_data

```

# snoRNA

The rest of the sequencing data didn't have raw counts.The CPM data provided will be read in for the snoRNA, snoRNA, snRNA, tRNA.

## Plasma:

### Read data in

```{r}

# CPM normalised plasma snoRNA data
plasma_snoRNA_data <- read.csv("./data/snoRNA_plasma.csv", row.names = 1)
colnames(plasma_snoRNA_data) <- gsub(pattern = "X", replacement = "Sample_", x = colnames(plasma_snoRNA_data))

```

Process raw data and make sure its in the same order as the phenotype table:

```{r}

plasma_snoRNA_data <- plasma_snoRNA_data[,match(plasma_pheno$Sample.Number, colnames(plasma_snoRNA_data))]

```

Look at the distribution of the data provided...

```{r, message=FALSE}

bp_plasma_snoRNA_data <- gg_boxplot(plasma_snoRNA_data,
                                log.transform =T,
                                title="Plasma snoRNA",
                                xlab="Sample ID", ylab="Log2 CPM")

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

bp_plasma_snoRNA_data

```

## Synovial fluid:

### Read data in

```{r}

# CPM normalised SF snoRNA data
sf_snoRNA_data <- read.csv("./data/snoRNA_sf.csv", row.names = 1)
colnames(sf_snoRNA_data) <- gsub(pattern = "X", replacement = "Sample_", x = colnames(sf_snoRNA_data))

```

Process raw data and make sure its in the same order as the phenotype table:

```{r}

sf_snoRNA_data <- sf_snoRNA_data[,match(sf_pheno$Sample.Number, colnames(sf_snoRNA_data))]

```

Look at the distribution of the data provided...

```{r, message=FALSE}

bp_sf_snoRNA_data <- gg_boxplot(sf_snoRNA_data,
                               log.transform=T,
                               title="Synovial fluid snoRNA",
                               xlab="Sample ID", ylab="Log2 CPM")

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

bp_sf_snoRNA_data

```

# snRNA

The rest of the sequencing data didn't have raw counts.The CPM data provided will be read in for the snRNA, snRNA, snRNA, tRNA.

## Plasma:

### Read data in

```{r}

# CPM normalised plasma snRNA data
plasma_snRNA_data <- read.csv("./data/snRNA_plasma.csv", row.names = 1)
colnames(plasma_snRNA_data) <- gsub(pattern = "X", replacement = "Sample_", x = colnames(plasma_snRNA_data))

```

Process raw data and make sure its in the same order as the phenotype table:

```{r}

plasma_snRNA_data <- plasma_snRNA_data[,match(plasma_pheno$Sample.Number, colnames(plasma_snRNA_data))]

```

Look at the distribution of the data provided...

```{r, message=FALSE}

bp_plasma_snRNA_data <- gg_boxplot(plasma_snRNA_data,
                                log.transform =T,
                                title="Plasma snRNA",
                                xlab="Sample ID", ylab="Log2 CPM")

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

bp_plasma_snRNA_data

```

## Synovial fluid:

### Read data in

```{r}

# CPM normalised SF snRNA data
sf_snRNA_data <- read.csv("./data/snRNA_sf.csv", row.names = 1)
colnames(sf_snRNA_data) <- gsub(pattern = "X", replacement = "Sample_", x = colnames(sf_snRNA_data))

```

Process raw data and make sure its in the same order as the phenotype table:

```{r}

sf_snRNA_data <- sf_snRNA_data[,match(sf_pheno$Sample.Number, colnames(sf_snRNA_data))]

```

Look at the distribution of the data provided...

```{r, message=FALSE}

bp_sf_snRNA_data <- gg_boxplot(sf_snRNA_data,
                               log.transform=T,
                               title="Synovial fluid snRNA",
                               xlab="Sample ID", ylab="Log2 CPM")

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

bp_sf_snRNA_data

```

# tRNA

The rest of the sequencing data didn't have raw counts.The CPM data provided will be read in for the tRNA, tRNA, tRNA, tRNA.

## Plasma:

### Read data in

```{r}

# CPM normalised plasma tRNA data
plasma_tRNA_data <- read.csv("./data/tRNA_plasma.csv", row.names = 1)
colnames(plasma_tRNA_data) <- gsub(pattern = "X", replacement = "Sample_", x = colnames(plasma_tRNA_data))

```

Process raw data and make sure its in the same order as the phenotype table:

```{r}

plasma_tRNA_data <- plasma_tRNA_data[,match(plasma_pheno$Sample.Number, colnames(plasma_tRNA_data))]

```

Look at the distribution of the data provided...

```{r, message=FALSE}

bp_plasma_tRNA_data <- gg_boxplot(plasma_tRNA_data,
                                log.transform =T,
                                title="Plasma tRNA",
                                xlab="Sample ID", ylab="Log2 CPM")

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

bp_plasma_tRNA_data

```


# Additional QC - missing values

## PCA

A problem with this analysis is the high number of zeroes present in the data. These could be biological zeroes or sampling zeroes. This problem has been reviewed in depth [elsewhere](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7568192/) [@silverman2020].


It is normal in sequencing analysis to add a small constant to the count data before log2 transforming it. In the case of sampling zeroes adding a small constant to all the data before log-transforming would be inappropriate. The plasma miRNA data will be used to demonstrate this problem below.

Create two dataframes, one a log transformed matrix with all the zeroes removed, the other a matrix with a small constant added before log transformation. 

```{r}

# Log2 transform data then remove the Inf values resulting from log2 transforming 0
# Inf2NA is a custom helper function
plasma_miRNA_data_log2 <- plasma_miRNA_data %>% log2(.) %>% Inf2NA(.)


# Add a small constant to the data then log transform
plasma_miRNA_data_log2_constant <- log2(plasma_miRNA_data+1)

```


PCAtools requires the row names in the metadata to be the same as the column names.

```{r}

row.names(plasma_pheno) <- plasma_pheno$Sample.Number

```


Carry out PCA on the log2 transformed data and the log2+1 transformed data:

```{r}

# Log2
plasma_miRNA_data_log2_PCA <- pca(na.omit(plasma_miRNA_data_log2), metadata = plasma_pheno, scale = TRUE, center = TRUE)

# Log2+1
plasma_miRNA_data_log2_constant_PCA <- pca(na.omit(plasma_miRNA_data_log2_constant), metadata = plasma_pheno, scale = TRUE, center = TRUE)

```


Now create PCA plots and correlation plots for the data. 

```{r, message=FALSE, warning=FALSE}

# Log2

# Correlation plot
log2_corplot <- eigencorplotPCA(plasma_miRNA_data_log2_PCA,
    metavars = c("Day", "Horse", "Date", "miRNA")) 


# PCA plot
log2_PCAplot <- plotPCA(plasma_miRNA_data_log2_PCA,
        PCs=c(1,2),
        colour.data=plasma_pheno$Day,
        shape.data=plasma_pheno$Horse,
        colour.lab="Day",
        shape.lab="Horse")

```

```{r, message=FALSE, warning=FALSE}

# Log2 + 1

# Correlation plot
log2_constant_corplot <- eigencorplotPCA(plasma_miRNA_data_log2_constant_PCA,
    metavars = c("Day", "Horse", "Date", "miRNA")) 


# PCA plot
log2_constant_PCAplot <- plotPCA(plasma_miRNA_data_log2_constant_PCA,
        PCs=c(1,2),
        colour.data=plasma_pheno$Day,
        shape.data=plasma_pheno$Horse,
        colour.lab="Day",
        shape.lab="Horse")

```


```{r, echo=FALSE, fig.height = 13, fig.width = 9, fig.align="center"}

ggarrange(ggarrange(log2_PCAplot, log2_constant_PCAplot, nrow = 1, ncol = 2, labels = c("A", "B")), log2_corplot, log2_constant_corplot,
          nrow = 3, 
          ncol = 1,
          labels = c("", "C", "D"))

```


## Pearson's correlation

Pearson's correlation of small RNA sequencing depth vs number of missing values:

```{r}

cor.test(colSums(isZero(plasma_miRNA_data)), plasma_pheno$miRNA, method = "pearson")

```


### Plasma 

```{r, echo=FALSE, message=FALSE, fig.height = 9, fig.width = 7, fig.align="center"}

# miRNA
plasma_miRNA_scatter <- depth_by_missing_scatter(plasma_miRNA_data,
                                          read.count=plasma_pheno$miRNA,
                                          x.lab="# missing plasma miRNA values",
                                          y.lab="# plasma miRNA reads")


# lncRNA
plasma_lncRNA_scatter <- depth_by_missing_scatter(plasma_lncRNA_data,
                                                 read.count=plasma_pheno$lncRNA,
                                                 x.lab="# missing plasma lncRNA values",
                                                 y.lab="# plasma lncRNA reads")


# snoRNA
plasma_snoRNA_scatter <- depth_by_missing_scatter(plasma_snoRNA_data,
                                                 read.count=plasma_pheno$snoRNA,
                                                 x.lab="# missing plasma snoRNA values",
                                                 y.lab="# plasma snoRNA reads")

# snRNA
plasma_snRNA_scatter <- depth_by_missing_scatter(plasma_snRNA_data,
                                                 read.count=plasma_pheno$snRNA,
                                                 x.lab="# missing plasma snRNA values",
                                                 y.lab="# plasma snRNA reads")

# tRNA
plasma_tRNA_scatter <- depth_by_missing_scatter(plasma_tRNA_data,
                                                 read.count=plasma_pheno$tRNA,
                                                 x.lab="# missing plasma tRNA values",
                                                 y.lab="# plasma tRNA reads")


ggarrange(plasma_miRNA_scatter, plasma_lncRNA_scatter, plasma_snoRNA_scatter, plasma_snRNA_scatter, plasma_tRNA_scatter,
          nrow = 3, 
          ncol = 2,
          labels = c("A", "B", "C", "D", "E"))

```




### Synovial fluid

```{r, echo=FALSE, message=FALSE, fig.height = 6, fig.width = 7, fig.align="center"}

# miRNA
sf_miRNA_scatter <- depth_by_missing_scatter(sf_miRNA_data,
                                                 read.count=sf_pheno$miRNA,
                                                 x.lab="# missing sf miRNA values",
                                                 y.lab="# sf miRNA reads")


# lncRNA
sf_lncRNA_scatter <- depth_by_missing_scatter(sf_lncRNA_data,
                                                  read.count=sf_pheno$lncRNA,
                                                  x.lab="# missing sf lncRNA values",
                                                  y.lab="# sf lncRNA reads")


# snoRNA
sf_snoRNA_scatter <- depth_by_missing_scatter(sf_snoRNA_data,
                                                  read.count=sf_pheno$snoRNA,
                                                  x.lab="# missing sf snoRNA values",
                                                  y.lab="# sf snoRNA reads")

# snRNA
sf_snRNA_scatter <- depth_by_missing_scatter(sf_snRNA_data,
                                                 read.count=sf_pheno$snRNA,
                                                 x.lab="# missing sf snRNA values",
                                                 y.lab="# sf snRNA reads")


ggarrange(sf_miRNA_scatter, sf_lncRNA_scatter, sf_snoRNA_scatter, sf_snRNA_scatter,
          nrow = 2, 
          ncol = 2,
          labels = c("A", "B", "C", "D"))

```



## Filter data by missingness 

```{r}

# Plasma
plasma_miRNA_data <- plasma_miRNA_data[rowSums(plasma_miRNA_data>0)>=round(ncol(plasma_miRNA_data)*0.6),]
plasma_lncRNA_data <- plasma_lncRNA_data[rowSums(plasma_lncRNA_data>0)>=round(ncol(plasma_lncRNA_data)*0.6),]
plasma_snoRNA_data <- plasma_snoRNA_data[rowSums(plasma_snoRNA_data>0)>=round(ncol(plasma_snoRNA_data)*0.6),]
plasma_snRNA_data <- plasma_snRNA_data[rowSums(plasma_snRNA_data>0)>=round(ncol(plasma_snRNA_data)*0.6),]
plasma_tRNA_data <- plasma_tRNA_data[rowSums(plasma_tRNA_data>0)>=round(ncol(plasma_tRNA_data)*0.6),]

# SF
sf_miRNA_data <- sf_miRNA_data[rowSums(sf_miRNA_data>0)>=round(ncol(sf_miRNA_data)*0.6),]
sf_lncRNA_data <- sf_lncRNA_data[rowSums(sf_lncRNA_data>0)>=round(ncol(sf_lncRNA_data)*0.6),]
sf_snoRNA_data <- sf_snoRNA_data[rowSums(sf_snoRNA_data>0)>=round(ncol(sf_snoRNA_data)*0.6),]
sf_snRNA_data <- sf_snRNA_data[rowSums(sf_snRNA_data>0)>=round(ncol(sf_snRNA_data)*0.6),]

```


# Write R data object outputs

Now that data has been filtered for sampling bias, log2 normalise and add constant to filtered data: 

```{r}

# Plasma 
saveRDS(plasma_pheno, "./data/plasma_sequencing_pheno.rds")
saveRDS(log2(plasma_miRNA_data+1), "./data/plasma_miRNA_data.rds")
saveRDS(log2(plasma_lncRNA_data+1), "./data/plasma_lncRNA_data.rds")
saveRDS(log2(plasma_snoRNA_data+1), "./data/plasma_snoRNA_data.rds")
saveRDS(log2(plasma_snRNA_data+1), "./data/plasma_snRNA_data.rds")
saveRDS(log2(plasma_tRNA_data+1), "./data/plasma_tRNA_data.rds")

# SF
saveRDS(sf_pheno, "./data/sf_sequencing_pheno.rds")
saveRDS(log2(sf_miRNA_data+1), "./data/sf_miRNA_data.rds")
saveRDS(log2(sf_lncRNA_data+1), "./data/sf_lncRNA_data.rds")
saveRDS(log2(sf_snoRNA_data+1), "./data/sf_snoRNA_data.rds")
saveRDS(log2(sf_snRNA_data+1), "./data/sf_snRNA_data.rds")


```



# References
