---
title: "Exploratory data analysis - sequencing data"
author: "ejjohnson93"
date: "2022-09-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
bibliography: references.bib
---

# Load libraries

```{r, message=FALSE, warning=FALSE}

source("./code/utility-functions.R") # load helper functions 
library(tidyverse)
library(cowplot) # plot theme
library(mixOmics) # multi-level PCA
library(viridis) # colour schemes
library(ggpubr) # utilities to arrange/annotate ggplot2 plots
library(PCAtools) # pca & correlation plot

```

# Load data

```{r}

# Plasma 
plasma_pheno_seq <- readRDS("./data/plasma_sequencing_pheno.rds") 
plasma_miRNA_data <- readRDS("./data/plasma_miRNA_data.rds") 
plasma_lncRNA_data <- readRDS("./data/plasma_lncRNA_data.rds")
plasma_snoRNA_data <- readRDS("./data/plasma_snoRNA_data.rds")
plasma_snRNA_data <- readRDS("./data/plasma_snRNA_data.rds")
plasma_tRNA_data <- readRDS("./data/plasma_tRNA_data.rds")

# SF
sf_pheno_seq <- readRDS("./data/sf_sequencing_pheno.rds")
sf_miRNA_data <- readRDS("./data/sf_miRNA_data.rds")
sf_lncRNA_data <- readRDS("./data/sf_lncRNA_data.rds")
sf_snoRNA_data <- readRDS("./data/sf_snoRNA_data.rds")
sf_snRNA_data <- readRDS("./data/sf_snRNA_data.rds")


```


# miRNA

## PCA

### Plasma


```{r}

row.names(plasma_pheno_seq) <- plasma_pheno_seq$Sample.Number


```


```{r}

miRNA_plasma_pca_res <- pca(na.omit(plasma_miRNA_data), metadata = plasma_pheno_seq, scale = TRUE, center = TRUE)


```


```{r, message=FALSE, warning=FALSE}

# PCA plot
plasma_miRNA_PCA <- plotPCA(miRNA_plasma_pca_res,
        PCs=c(1,2),
        colour.data=plasma_pheno_seq$Day,
        shape.data=plasma_pheno_seq$Horse,
        colour.lab="Day",
        shape.lab="Horse")

plasma_miRNA_PCA


# Correlation plot
plasma_miRNA_corplot <- eigencorplotPCA(miRNA_plasma_pca_res,
    metavars = c("Day", "Horse", "Date", "miRNA")) 


plasma_miRNA_corplot

```


### SF


```{r}

row.names(sf_pheno_seq) <- sf_pheno_seq$Sample.Number


```


```{r}

miRNA_sf_pca_res <- pca(na.omit(sf_miRNA_data), metadata = sf_pheno_seq, scale = TRUE, center = TRUE)


```


```{r, message=FALSE, warning=FALSE}

# PCA plot
sf_miRNA_PCA <- plotPCA(miRNA_sf_pca_res,
                            PCs=c(1,2),
                            colour.data=sf_pheno_seq$Day,
                            shape.data=sf_pheno_seq$Horse,
                            colour.lab="Day",
                            shape.lab="Horse")

sf_miRNA_PCA


# Correlation plot
sf_miRNA_corplot <- eigencorplotPCA(miRNA_sf_pca_res,
                                        metavars = c("Day", "Horse", "Date", "miRNA", "Group")) 


sf_miRNA_corplot

```


## Multi-level PCA

### Plasma

```{r, fig.show="hide", message=FALSE}

plasma_miRNA_ml_PCA <- multi_level_pca(na.omit(plasma_miRNA_data),
                multilevel = plasma_pheno_seq$Horse,
                group = plasma_pheno_seq$Day, group.lab = "Day", title = "miRNA - Plasma multi-level PCA")


```

```{r}

plasma_miRNA_ml_PCA

```


### SF

```{r, fig.show="hide", message=FALSE}

sf_miRNA_ml_PCA <- multi_level_pca(na.omit(sf_miRNA_data),
                multilevel = sf_pheno_seq$Horse,
                group = sf_pheno_seq$Day, group.lab = "Day", title = "miRNA - Synovial fluid multi-level PCA")


```

```{r}

sf_miRNA_ml_PCA

```



# lncRNA

## PCA

### Plasma


```{r}

lncRNA_plasma_pca_res <- pca(na.omit(plasma_lncRNA_data), metadata = plasma_pheno_seq, scale = TRUE, center = TRUE)


```


```{r, message=FALSE, warning=FALSE}

# PCA plot
plasma_lncRNA_PCA <- plotPCA(lncRNA_plasma_pca_res,
                            PCs=c(1,2),
                            colour.data=plasma_pheno_seq$Day,
                            shape.data=plasma_pheno_seq$Horse,
                            colour.lab="Day",
                            shape.lab="Horse")

plasma_lncRNA_PCA


# Correlation plot
plasma_lncRNA_corplot <- eigencorplotPCA(lncRNA_plasma_pca_res,
                                        metavars = c("Day", "Horse", "Date", "lncRNA")) 


plasma_lncRNA_corplot

```


### SF


```{r}

lncRNA_sf_pca_res <- pca(na.omit(sf_lncRNA_data), metadata = sf_pheno_seq, scale = TRUE, center = TRUE)


```


```{r, message=FALSE, warning=FALSE}

# PCA plot
sf_lncRNA_PCA <- plotPCA(lncRNA_sf_pca_res,
                        PCs=c(1,2),
                        colour.data=sf_pheno_seq$Day,
                        shape.data=sf_pheno_seq$Horse,
                        colour.lab="Day",
                        shape.lab="Horse")

sf_lncRNA_PCA


# Correlation plot
sf_lncRNA_corplot <- eigencorplotPCA(lncRNA_sf_pca_res,
                                    metavars = c("Day", "Horse", "Date", "lncRNA", "Group")) 


sf_lncRNA_corplot

```


## Multi-level PCA

### Plasma

```{r, fig.show="hide", message=FALSE}

plasma_lncRNA_ml_PCA <- multi_level_pca(na.omit(plasma_lncRNA_data),
                                       multilevel = plasma_pheno_seq$Horse,
                                       group = plasma_pheno_seq$Day, group.lab = "Day", title = "lncRNA - Plasma multi-level PCA")


```

```{r}

plasma_lncRNA_ml_PCA

```


### SF

```{r, fig.show="hide", message=FALSE}

sf_lncRNA_ml_PCA <- multi_level_pca(na.omit(sf_lncRNA_data),
                                   multilevel = sf_pheno_seq$Horse,
                                   group = sf_pheno_seq$Day, group.lab = "Day", title = "lncRNA - Synovial fluid multi-level PCA")

```

```{r}

sf_lncRNA_ml_PCA

```


# snRNA

## PCA

### Plasma


```{r}

snRNA_plasma_pca_res <- pca(na.omit(plasma_snRNA_data), metadata = plasma_pheno_seq, scale = TRUE, center = TRUE)


```


```{r, message=FALSE, warning=FALSE}

# PCA plot
plasma_snRNA_PCA <- plotPCA(snRNA_plasma_pca_res,
                             PCs=c(1,2),
                             colour.data=plasma_pheno_seq$Day,
                             shape.data=plasma_pheno_seq$Horse,
                             colour.lab="Day",
                             shape.lab="Horse")

plasma_snRNA_PCA


# Correlation plot
plasma_snRNA_corplot <- eigencorplotPCA(snRNA_plasma_pca_res,
                                         metavars = c("Day", "Horse", "Date", "snRNA")) 


plasma_snRNA_corplot

```


### SF


```{r}

snRNA_sf_pca_res <- pca(na.omit(sf_snRNA_data), metadata = sf_pheno_seq, scale = TRUE, center = TRUE)


```


```{r, message=FALSE, warning=FALSE}

# PCA plot
sf_snRNA_PCA <- plotPCA(snRNA_sf_pca_res,
                         PCs=c(1,2),
                         colour.data=sf_pheno_seq$Day,
                         shape.data=sf_pheno_seq$Horse,
                         colour.lab="Day",
                         shape.lab="Horse")

sf_snRNA_PCA


# Correlation plot
sf_snRNA_corplot <- eigencorplotPCA(snRNA_sf_pca_res,
                                     metavars = c("Day", "Horse", "Date", "snRNA", "Group")) 


sf_snRNA_corplot

```


## Multi-level PCA

### Plasma

```{r, fig.show="hide", message=FALSE}

plasma_snRNA_ml_PCA <- multi_level_pca(na.omit(plasma_snRNA_data),
                                        multilevel = plasma_pheno_seq$Horse,
                                        group = plasma_pheno_seq$Day, group.lab = "Day", title = "snRNA - Plasma multi-level PCA")


```

```{r}

plasma_snRNA_ml_PCA

```


### SF

```{r, fig.show="hide", message=FALSE}

sf_snRNA_ml_PCA <- multi_level_pca(na.omit(sf_snRNA_data),
                                    multilevel = sf_pheno_seq$Horse,
                                    group = sf_pheno_seq$Day, group.lab = "Day", title = "snRNA - Synovial fluid multi-level PCA")

```

```{r}

sf_snRNA_ml_PCA

```


# Conclusions

Will not include snoRNA data in clustering analysis due to extremely high proportion of missing values (only 3 rows in SF data and plasma data after filtering). 
Notable perhaps that URS00006ACE58_9796 and URS000068B4BE_9796 present in both.


