---
title: "Multi-omics - joint dimensionality reduction"
author: "Emily Johnson"
date: "2022-09-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Load libraries

```{r, message=FALSE, warning=FALSE}

source("./code/utility-functions.R") # load helper functions 
library(tidyverse)
library(MOFA2) # factor analysis
library(ggpubr) # utilities to arrange/annotate ggplot2 plots
library(cowplot) # plot theme
library(viridis) # colour schemes

```

# Load and process data

```{r}

# Plasma 
plasma_annotation <- readRDS("./data/plasma_proteomics_annotation.rds")
plasma_pheno <- readRDS("./data/plasma_pheno.rds") 
plasma_proteomics_data <- readRDS("./data/plasma_data_norm.rds")
plasma_miRNA_data <- readRDS("./data/plasma_miRNA_data.rds") 
plasma_lncRNA_data <- readRDS("./data/plasma_lncRNA_data.rds")
plasma_snRNA_data <- readRDS("./data/plasma_snRNA_data.rds")

# SF
sf_annotation <- readRDS("./data/sf_proteomics_annotation.rds")
sf_pheno <- readRDS("./data/sf_pheno.rds")
sf_proteomics_data <- readRDS("./data/sf_data_norm.rds")
sf_miRNA_data <- readRDS("./data/sf_miRNA_data.rds")
sf_lncRNA_data <- readRDS("./data/sf_lncRNA_data.rds")
sf_snRNA_data <- readRDS("./data/sf_snRNA_data.rds")


```

Convert 0s to NAs as previous analysis showed they were sampling zeroes rather than biological zeroes:

```{r}

# Plasma 
plasma_miRNA_data[plasma_miRNA_data == 0] <- NA
plasma_lncRNA_data[plasma_lncRNA_data == 0] <- NA
plasma_snRNA_data[plasma_snRNA_data == 0] <- NA

# SF
sf_miRNA_data[sf_miRNA_data == 0] <- NA
sf_lncRNA_data[sf_lncRNA_data == 0] <- NA
sf_snRNA_data[sf_snRNA_data == 0] <- NA


```

# MOFA

## Plasma

Visualuse the distributions of the data:

```{r, echo=FALSE, message=FALSE, warning=FALSE}

proteomics_hist <- gg_histogram(plasma_proteomics_data, ylab="Frequency", xlab="Log2 Intensity")
miRNA_hist <- gg_histogram(plasma_miRNA_data, ylab="Frequency", xlab="Log2 CPM")
lncRNA_hist <- gg_histogram(plasma_lncRNA_data, ylab="Frequency", xlab="Log2 CPM")
snRNA_hist <- gg_histogram(plasma_snRNA_data, ylab="Frequency", xlab="Log2 CPM")


ggarrange(proteomics_hist, miRNA_hist, lncRNA_hist, snRNA_hist)

```

MOFA2 needs a column in the phenotype data called 'sample'. Rename the 'Sample_ID' column to 'sample'.

```{r}

colnames(plasma_pheno)[5] <- "sample"

```

As an input MOFA2 will take a list containing all the different data. Create this list:

```{r}

plasma_mofa_dat <- list(proteomics = as.matrix(plasma_proteomics_data),
                 miRNA = as.matrix(plasma_miRNA_data),
                 lncRNA = as.matrix(plasma_lncRNA_data),
                 snRNA = as.matrix(plasma_snRNA_data))

```

Create MOFAobject:

```{r}

MOFAobject <- create_mofa(plasma_mofa_dat)
print(MOFAobject)

```

The number of features per view can be visualised with `plot_data_overview`. Missing samples would be represented as gaps in the block colours.

```{r}

plot_data_overview(MOFAobject)

```

Set model parameters. Here three objects are created containing the data options, model options and training options. If the raw counts were used a poisson distribution would be more appropriate.

```{r}

# Scale views
data_opts <- get_default_data_options(MOFAobject)
data_opts$scale_views <- TRUE
head(data_opts)


# Set number of factors
model_opts <- get_default_model_options(MOFAobject)
model_opts$num_factors <- 6
#model_opts$likelihoods[2:4] <- "poisson"
head(model_opts)


# Set training options
train_opts <- get_default_training_options(MOFAobject)
train_opts$convergence_mode <- "slow"
train_opts$seed <- 123
train_opts$maxiter <- 3000
head(train_opts)

```

Prepare MOFA object with modified parameters:

```{r}

MOFAobject <- prepare_mofa(
  object = MOFAobject,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)

```

Train model:

```{r, message=FALSE, warning=FALSE}

outfile <- file.path(getwd(),"plasma_model.hdf5")
MOFAobject.trained <- run_mofa(MOFAobject, outfile, use_basilisk=TRUE)

```

Add phenotype data to the model:

```{r}

samples_metadata(MOFAobject.trained) <- plasma_pheno

```

Visualise variance explained:

```{r}

plot_variance_explained(
  MOFAobject.trained, 
  x = "view", 
  y = "factor", 
  plot_total = T
)

```

Majority is from proteomics data - likely due to the higher number of features.

Plot correlation between factors:

```{r}

plot_factor_cor(MOFAobject.trained)

```

Correlation between factors and metadata:

```{r, message=FALSE, warning=FALSE}

correlate_factors_with_covariates(MOFAobject.trained, 
  covariates = c("Day", "Horse"), 
  plot="log_pval", 
  alpha="0.15" # p-value threshold
)


correlate_factors_with_covariates(MOFAobject.trained, 
  covariates = c("Day", "Horse"), 
  plot="r"
)


```

Factor 2 is significantly correlated with "day". Factors 5 and 6 are correlated with horse. Most of the variance in factor 2 is explained by the proteomics data. The variance in factor 1 mostly comes from the miRNA and lncRNA data, however that doesn't appear to correspond to a biological effect. Factor 3 shows slight correlation with "day". As the variance is explained by the proteomics and snRNA data this might also be interesting to explore.

We can inspect the weights of the different features contributing to each factor.

```{r}

proteomics_f2 <- plot_top_weights(MOFAobject.trained, 
                  view = "proteomics", 
                  factor = 2, 
                  nfeatures = 10,
                  scale = F, 
                  abs = T
)

# Annotate so Uniprot accessions are replaced by protein names
# The accessions are stored as factors so need to be converted back into character vectors
proteomics_f2$data$feature_id <- plasma_annotation[as.character(proteomics_f2$data$feature_id),]

```

```{r, echo=FALSE}

proteomics_f2

```

Plot weights for factor 2:

```{r}

plot_weights(MOFAobject.trained,
 view = "proteomics",
 factor = 2,
 nfeatures = 50,     # Top number of features to highlight
 scale = F           # Scale weights from -1 to 1
)


plot_weights(MOFAobject.trained,
 view = "miRNA",
 factor = 2,
 nfeatures = 25,     # Top number of features to highlight
 scale = F           # Scale weights from -1 to 1
)


```

The distribution of weights for the top ranked proteins contributing towards factor 2 are heavily negatively skewed. Only Q28377, F7DZE7, A0A5F5PEG5, and A0A5F5PWX5 are positively associated in the top 50. If the weights for miRNA are plotted it can be seen their contribution to factor 2 is low overall, as the weights range from -0.001 to 0.001.

If we plot factor 1 instead, the weight of the miRNAs is much higher.

```{r}

miRNA_f2 <- plot_top_weights(MOFAobject.trained, 
                  view = "miRNA", 
                  factor = 2, 
                  nfeatures = 10,
                  scale = F, 
                  abs = T
)

miRNA_f1 <- plot_top_weights(MOFAobject.trained, 
                  view = "miRNA", 
                  factor = 1, 
                  nfeatures = 10,
                  scale = F, 
                  abs = T
)

```

```{r, echo=FALSE}

ggarrange(miRNA_f2, miRNA_f1)

```

It is interesting that miR-126-5P and miR-126-3P are among the miRNAs contributing to factor 2. <https://pubmed.ncbi.nlm.nih.gov/33627637/>

Visualise the weights for the proteomics and snRNA data contributing to factor 3:

```{r}

snRNA_f3 <- plot_top_weights(MOFAobject.trained, 
                  view = "snRNA", 
                  factor = 3, 
                  nfeatures = 10,
                  scale = F, 
                  abs = T
)


snRNA_f4 <- plot_top_weights(MOFAobject.trained, 
                  view = "snRNA", 
                  factor = 4, 
                  nfeatures = 10,
                  scale = F, 
                  abs = T
)

```

```{r, echo=FALSE}

ggarrange(snRNA_f3, snRNA_f4)

```

```{r}

proteomics_f3 <- plot_top_weights(MOFAobject.trained, 
                  view = "proteomics", 
                  factor = 3, 
                  nfeatures = 10,
                  scale = F, 
                  abs = T
)

# Annotate so Uniprot accessions are replaced by protein names
# The accessions are stored as factors so need to be converted back into character vectors
proteomics_f3$data$feature_id <- plasma_annotation[as.character(proteomics_f3$data$feature_id),]

```

```{r, echo=FALSE}

proteomics_f3

```

Create heatmaps for proteomics and miRNA data (factor 2):

```{r}

plot_data_heatmap(MOFAobject.trained, 
  view = "proteomics",
  factor = 2,  
  features = 25,
  cluster_rows = TRUE, cluster_cols = TRUE,
  show_rownames = TRUE, show_colnames = TRUE,
  scale = "row",
  denoise = TRUE,
  color = viridis(10),
  border_color = NA
)

```

```{r}

plot_data_heatmap(MOFAobject.trained, 
  view = "miRNA",
  factor = 2,  
  features = 25,
  cluster_rows = TRUE, cluster_cols = TRUE,
  show_rownames = TRUE, show_colnames = TRUE,
  scale = "row",
  denoise = TRUE,
  color = viridis(10),
  border_color = NA
)

```

Plot factors 2 and 3 together (equivalent to PCA):

```{r}

p <- plot_factors(MOFAobject.trained, 
  factors = c(2,3), 
  color_by = "Day",
  shape_by = "Horse",
  dot_size = 4,
  show_missing = T
)

p + theme_cowplot()

```

Project the samples onto factor 2 and 3 for each time point.

```{r}

# Add day as a factor to metadata for visualisation purposes
MOFAobject.trained@samples_metadata$day_group <- as.factor(plasma_pheno$Day)

plot_factor(MOFAobject.trained, 
  factors = 2, 
  color_by = "day_group",
  add_violin = TRUE,
  dodge = TRUE
)


plot_factor(MOFAobject.trained, 
  factors = 3, 
  color_by = "day_group",
  add_violin = TRUE,
  dodge = TRUE
)

```


Save results:

```{r}

plasma_res <- get_weights(
    MOFAobject.trained,
    views = "all",
    factors = "all",
    abs = FALSE,
    scale = FALSE,
    as.data.frame = FALSE
)



plasma_res_df <- get_weights(
    MOFAobject.trained,
    views = "all",
    factors = "all",
    abs = FALSE,
    scale = FALSE,
    as.data.frame = TRUE
)


```

```{r}

saveRDS(plasma_res, "./data/mofa2_plasma.rds")
saveRDS(plasma_res_df, "./data/mofa2_plasma_df.rds")

```

## Synovial fluid

```{r, echo=FALSE, message=FALSE, warning=FALSE}

proteomics_hist <- gg_histogram(sf_proteomics_data, ylab="Frequency", xlab="Log2 Intensity")
miRNA_hist <- gg_histogram(sf_miRNA_data, ylab="Frequency", xlab="Log2 CPM")
lncRNA_hist <- gg_histogram(sf_lncRNA_data, ylab="Frequency", xlab="Log2 CPM")
snRNA_hist <- gg_histogram(sf_snRNA_data, ylab="Frequency", xlab="Log2 CPM")


ggarrange(proteomics_hist, miRNA_hist, lncRNA_hist, snRNA_hist)

```

```{r}

colnames(sf_pheno)[6] <- "sample"

```

```{r}

sf_mofa_dat <- list(proteomics = as.matrix(sf_proteomics_data),
                        miRNA = as.matrix(sf_miRNA_data),
                        lncRNA = as.matrix(sf_lncRNA_data),
                        snRNA = as.matrix(sf_snRNA_data))

```

Create MOFAojbect:

```{r}

MOFAobject <- create_mofa(sf_mofa_dat)
print(MOFAobject)

```

```{r}

plot_data_overview(MOFAobject)

```

Modify model parameters:

```{r}

# Scale views
data_opts <- get_default_data_options(MOFAobject)
data_opts$scale_views <- TRUE
head(data_opts)


# Set number of factors
model_opts <- get_default_model_options(MOFAobject)
model_opts$num_factors <- 6
head(model_opts)


# Set training options
train_opts <- get_default_training_options(MOFAobject)
train_opts$convergence_mode <- "slow"
train_opts$seed <- 123
train_opts$maxiter <- 3000
head(train_opts)

```

Prepare MOFA object with modified parameters:

```{r}

MOFAobject <- prepare_mofa(
  object = MOFAobject,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)

```

Train model:

```{r, message=FALSE, warning=FALSE}

outfile <- file.path(getwd(),"sf_model.hdf5")
MOFAobject.trained <- run_mofa(MOFAobject, outfile, use_basilisk=TRUE)

```

Add phenotype data to the model:

```{r}

samples_metadata(MOFAobject.trained) <- sf_pheno

# Add a group-by-day factor 
MOFAobject.trained@samples_metadata$Group_day <- as.factor(paste0(sf_pheno$Group, "_", sf_pheno$Day))

```

Visualise variance explained:

```{r}

plot_variance_explained(
  MOFAobject.trained, 
  x = "view", 
  y = "factor", 
  plot_total = T
)


```

Majority is from proteomics data - likely due to the higher number of features.

Plot correlation between factors:

```{r}

plot_factor_cor(MOFAobject.trained)

```

Correlation between factors and metadata:

```{r, message=FALSE, warning=FALSE}

correlate_factors_with_covariates(MOFAobject.trained, 
                                  covariates = c("Day", "Horse", "Group", "Group_day"), 
                                  plot="log_pval",
                                  alpha="0.05" # p-value threshold
)


correlate_factors_with_covariates(MOFAobject.trained, 
  covariates = c("Day", "Horse", "Group", "Group_day"), 
  plot="r"
)

```

Factor 4 is significantly correlated with day, though factors 2-5 seem to have some correlation with day. Factor 4 is mostly explained by variance in the miRNA. Factor 2 is correlated with the 'group_day' variable, though its worth noting this factor had to be coerced into a numeric variable. For factor 2 miRNA and lncRNA explain more of the variance.

It may be worth looking at the proteomics weight for factor 1 and 5. In factor 1's case the variance is most strongly correlated with the proteomics but this may be because there is a much higher number of features. Factor 5 seems to have a slight correlation with the biological factors of interest.

```{r}

miRNA_f2 <- plot_top_weights(MOFAobject.trained, 
                  view = "miRNA", 
                  factor = 2, 
                  nfeatures = 10,
                  scale = F, 
                  abs = T
)

```

```{r}

lncRNA_f2 <- plot_top_weights(MOFAobject.trained, 
                  view = "lncRNA", 
                  factor = 2, 
                  nfeatures = 10,
                  scale = F, 
                  abs = T
)

```

```{r}

miRNA_f4 <- plot_top_weights(MOFAobject.trained, 
                  view = "miRNA", 
                  factor = 4, 
                  nfeatures = 10,
                  scale = F, 
                  abs = T
)

```

```{r}

proteomics_f3 <- plot_top_weights(MOFAobject.trained, 
                  view = "proteomics", 
                  factor = 3, 
                  nfeatures = 10,
                  scale = F, 
                  abs = T
)


# Annotate so Uniprot accessions are replaced by protein names
# The accessions are stored as factors so need to be converted back into character vectors
proteomics_f3$data$feature_id <- sf_annotation[as.character(proteomics_f3$data$feature_id),]

```

Plot miRNA factor 2, lncRNA factor 2, miRNA factor 4 and proteomics factor 3:

```{r, echo=FALSE}

ggarrange(miRNA_f2, lncRNA_f2, miRNA_f4)


proteomics_f3

```

Plot weights for factor 2:

```{r}

plot_weights(MOFAobject.trained,
 view = "miRNA",
 factor = 2,
 nfeatures = 20,     # Top number of features to highlight
 scale = F           # Scale weights from -1 to 1
)


plot_weights(MOFAobject.trained,
 view = "lncRNA",
 factor = 2,
 nfeatures = 20,     # Top number of features to highlight
 scale = F           # Scale weights from -1 to 1
)



```

```{r}

plot_data_heatmap(MOFAobject.trained, 
                  view = "miRNA",
                  factor = 4,  
                  features = 10,
                  cluster_rows = TRUE, cluster_cols = TRUE,
                  show_rownames = TRUE, show_colnames = TRUE,
                  scale = "row",
                  denoise = TRUE,
                  color = viridis(10),
                  border_color = NA
)

```

Project the samples onto factor 2 & 4 for each time point.

```{r}

# Add day as a factor to metadata for visualisation purposes
MOFAobject.trained@samples_metadata$group <- as.factor(paste0(sf_pheno$Group, "_", sf_pheno$Day))

```

```{r}

plot_factor(MOFAobject.trained, 
  factors = 2, 
  color_by = "group",
  add_violin = TRUE,
  dodge = TRUE
)

```

```{r}

plot_factor(MOFAobject.trained, 
  factors = 4, 
  color_by = "group",
  add_violin = TRUE,
  dodge = TRUE
)

```


Save results:

```{r}

sf_res <- get_weights(
    MOFAobject.trained,
    views = "all",
    factors = "all",
    abs = FALSE,
    scale = FALSE,
    as.data.frame = FALSE
)



sf_res_df <- get_weights(
    MOFAobject.trained,
    views = "all",
    factors = "all",
    abs = FALSE,
    scale = FALSE,
    as.data.frame = TRUE
)


```

```{r}

saveRDS(sf_res, "./data/mofa2_sf.rds")
saveRDS(sf_res_df, "./data/mofa2_sf_df.rds")

```

If more time was available would also compare/contrast JIVE and iCluster.
