---
title: "Multi-omics - clustering"
author: "Emily Johnson"
date: "2022-09-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Load libraries

```{r, message=FALSE, warning=FALSE}

library(tidyverse)
library(Mfuzz)
library(dtwclust)
library(cowplot) # ggplot2 theme

```

# Load and process data

```{r}

# Plasma 
plasma_annotation <- readRDS("./data/plasma_proteomics_annotation.rds")
plasma_pheno <- readRDS("./data/plasma_pheno.rds") 
plasma_proteomics_data <- readRDS("./data/plasma_data_norm.rds")
plasma_miRNA_data <- readRDS("./data/plasma_miRNA_data.rds") 
plasma_lncRNA_data <- readRDS("./data/plasma_lncRNA_data.rds")
plasma_snRNA_data <- readRDS("./data/plasma_snRNA_data.rds")

# SF
sf_annotation <- readRDS("./data/sf_proteomics_annotation.rds")
sf_pheno <- readRDS("./data/sf_pheno.rds")
sf_proteomics_data <- readRDS("./data/sf_data_norm.rds")
sf_miRNA_data <- readRDS("./data/sf_miRNA_data.rds")
sf_lncRNA_data <- readRDS("./data/sf_lncRNA_data.rds")
sf_snRNA_data <- readRDS("./data/sf_snRNA_data.rds")


```

Convert 0s to NAs as previous analysis showed they were sampling zeroes rather than biological zeroes: 

```{r}

# Plasma 
plasma_miRNA_data[plasma_miRNA_data == 0] <- NA
plasma_lncRNA_data[plasma_lncRNA_data == 0] <- NA
plasma_snRNA_data[plasma_snRNA_data == 0] <- NA

# SF
sf_miRNA_data[sf_miRNA_data == 0] <- NA
sf_lncRNA_data[sf_lncRNA_data == 0] <- NA
sf_snRNA_data[sf_snRNA_data == 0] <- NA


```

Remove universal spike-ins:

```{r}

plasma_miRNA_data <- plasma_miRNA_data[-c(1:3),]
sf_miRNA_data <- sf_miRNA_data[-c(1:3),]

```


## Variance filtering

Before Z-score normalisation we filter the data to only include features with high variance. This is because protein/sequencing data is noisy. Noisy features can then be assigned to clusters even if their expression pattern isn’t representative of the cluster.


```{r}

vars <- apply(plasma_proteomics_data, 1, var, na.rm=TRUE)
means <- apply(plasma_proteomics_data, 1, mean, na.rm=TRUE)

plasma_proteomics_data <- plasma_proteomics_data[which(vars > 1 & means >2),]
plasma_proteomics_data <- plasma_proteomics_data[which(rowMeans(is.na(plasma_proteomics_data)) < 0.2), ]

```


## Z-score normalisation

### Plasma

```{r}

# Proteomics
plasma_proteomics_Znorm <- t(apply(plasma_proteomics_data, 1, function(x) scale(x, center = T, scale = T)))
colnames(plasma_proteomics_Znorm) <- colnames(plasma_proteomics_data)

# miRNA
plasma_miRNA_Znorm <- t(apply(plasma_miRNA_data, 1, function(x) scale(x, center = T, scale = T)))
colnames(plasma_miRNA_Znorm) <- colnames(plasma_miRNA_data)


# lncRNA
plasma_lncRNA_Znorm <- t(apply(plasma_lncRNA_data, 1, function(x) scale(x, center = T, scale = T)))
colnames(plasma_lncRNA_Znorm) <- colnames(plasma_lncRNA_data)


# snRNA
plasma_snRNA_Znorm <- t(apply(plasma_snRNA_data, 1, function(x) scale(x, center = T, scale = T)))
colnames(plasma_snRNA_Znorm) <- colnames(plasma_snRNA_data)

```


Concatenate plasma data together:

```{r}

plasma_data <- rbind(plasma_proteomics_Znorm, plasma_miRNA_Znorm, plasma_lncRNA_Znorm, plasma_snRNA_Znorm)

```


### Synovial fluid

```{r}

# Proteomics
sf_proteomics_Znorm <- t(apply(sf_proteomics_data, 1, function(x) scale(x, center = T, scale = T)))
colnames(sf_proteomics_Znorm) <- colnames(sf_proteomics_data)

# miRNA
sf_miRNA_Znorm <- t(apply(sf_miRNA_data, 1, function(x) scale(x, center = T, scale = T)))
colnames(sf_miRNA_Znorm) <- colnames(sf_miRNA_data)


# lncRNA
sf_lncRNA_Znorm <- t(apply(sf_lncRNA_data, 1, function(x) scale(x, center = T, scale = T)))
colnames(sf_lncRNA_Znorm) <- colnames(sf_lncRNA_data)


# snRNA
sf_snRNA_Znorm <- t(apply(sf_snRNA_data, 1, function(x) scale(x, center = T, scale = T)))
colnames(sf_snRNA_Znorm) <- colnames(sf_snRNA_data)

```


Concatenate sf data together:
  
```{r}

sf_data <- rbind(sf_proteomics_Znorm, sf_miRNA_Znorm, sf_lncRNA_Znorm, sf_snRNA_Znorm)

```



## Collapse replicates and re-order by time

Plasma:

```{r}

plasma_time_df <- data.frame("Day_0" = rowMeans(plasma_data[,plasma_pheno$Day == "0"], na.rm=TRUE),
         "Day_10" = rowMeans(plasma_data[,plasma_pheno$Day == "10"], na.rm=TRUE),
         "Day_35" = rowMeans(plasma_data[,plasma_pheno$Day == "35"], na.rm=TRUE),
         "Day_42" = rowMeans(plasma_data[,plasma_pheno$Day == "42"], na.rm=TRUE),
         "Day_49" = rowMeans(plasma_data[,plasma_pheno$Day == "49"], na.rm=TRUE),
         "Day_56" = rowMeans(plasma_data[,plasma_pheno$Day == "56"], na.rm=TRUE),
         "Day_63" = rowMeans(plasma_data[,plasma_pheno$Day == "63"], na.rm=TRUE))


```



# Mfuzz

```{r}

expData <- ExpressionSet(as.matrix(na.omit(plasma_time_df)))

```

```{r}

m1 <- mestimate(expData)

```

```{r}

cl <- mfuzz(expData,c=6,m=m1)

```

```{r}

mfuzz.plot(eset = expData, cl, mfrow = c(2,3))

```


# dtwclust

```{r, fig.show='hide', message=FALSE}

pc <- tsclust(na.omit(plasma_time_df), type = "partitional", k = 6L, 
              distance = "dtw", centroid = "pam", 
              seed = 3247L, trace = TRUE,
              args = tsclust_args(dist = list(window.size = 20L)))



time=gsub("Day_", "", colnames(plasma_time_df))

pc_plot <- plot(pc, time=time) + 
            theme_cowplot() +
            xlab("Time (days") +
            ylab("Z-score") 

```

```{r, echo=FALSE}

pc_plot 

```

```{r, fig.show='hide', message=FALSE}

fc <- tsclust(na.omit(plasma_time_df), type = "fuzzy", k = 6L,
              distance = "L2",
              seed = 123L)


fc_plot <- plot(fc, time=time) + 
            theme_cowplot() +
            xlab("Time (days") +
            ylab("Z-score") 


```

```{r, echo=FALSE}

fc_plot 

```

https://stackoverflow.com/questions/42197984/how-to-get-the-result-of-dtwclust
